A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-04-16 paper.Rmd
switched facetting in Fig 17 to match the previous two figures

---
title: |
  A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol
type: ARTICLE TEMPLATE
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a, b, c
    email: emi.tanaka@anu.edu.au
  - name: Susan VanderPlas
    affil: d
    email: susan.vanderplas@unl.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Biological Data Science Institute, Australian National University, Acton, ACT, Australia
  - num: c
    address: |
      Research School of Finance, Actuarial Studies and Statistics, Australian National University, Acton, ACT, Australia
  - num: d
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: paper.bib
abstract: |
  Regression experts consistently recommend plotting residuals for model diagnosis, despite the existence of numerous hypothesis test procedures. This paper provides evidence for why this is good advice, using data from a visual inference experiment. We show how conventional tests are too sensitive to be practically useful, and how the lineup protocol can be used to yield reliable and consistent reading of residual plots for better practical application. 
keywords: |
  statistical graphics; data visualization; visual inference; hypothesis testing; reression analysis; cognitive perception; simulation; practical significance; effect size
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@

```{r different-x-dist-poly-power, fig.pos="t!", fig.height = 6, fig.cap = "Comparison of power on lineups with different fitted value distributions for conventional and visual tests for non-linearity and heteroskedasticity patterns. The power curves of conventional tests for non-linearity and heteroskedasticity patterns are produced by RESET tests and BP tests, respectively. Power curves of visual tests are estimated using five evaluations on each lineup. For lineups with a uniform fitted value distribution, the five evaluations are repeatedly sampled from the total eleven evaluations to give multiple power curves (grey). Surprisingly, the fitted value distribution has produces more variability in the power of conventional tests than visual tests. Uneven distributions, normal and lognormal distributions, tend to yield lower power.", dev = 'png', dpi = 450}

@@

  # Theme
  theme_light() +
  scale_color_manual(values = rcartocolor::carto_pal(4, "Safe")[c(4,1,3,2)]) +
  xlab(expression(log[e] (Effect_size))) +
  ylab("Power") +
  labs(col = "", linetype = "Test") +
  facet_grid(conv_or_not~type, scales = "free_x") +
  theme(legend.position = "bottom")
```

Figure \ref{fig:different-x-dist-poly-power} examines the impact of the fitted value distribution on the power of conventional (top row) and visual (bottom row) tests for both the non-linearity (left column) and heteroskedasticity (right column) patterns. For conventional tests, only the power curves of appropriate tests which are RESET tests and BP tests for non-linearity pattern and heteroskedasticity pattern respectively are shown. For visual tests, note that, more evaluations on lineup with uniform fitted value distribution were collected, so to have a fair comparison, we randomly sample five from the total eleven evaluations to estimate the power curves. Re-sampling produces the multiple curves for the uniform, and provides an indication of the variation of the power.

@@

Motivated by the advice of regression analysis experts, that residual plots as opposed to conventional tests are an indispensible methods for assessing model fit, a human subjects experiment was conducted using visual inference. The experiment tested two primary departures from good residuals: non-linearity and heteroskedasticity.

The experiment found that conventional residual-based statistical tests are more sensitive to weak residual departures from model assumptions than visual tests as would be evaluated by humans. That is, conventional tests conclude there are problems with the model fit almost twice as often as humans would. They often reject when departures in the form of non-linearity and heteroskedasticity are not visible to a human.

One might say that this is correct behaviour, but it can be argued that the conventional tests are rejecting when it is not necessary. Many of these rejections happen even when downstream analysis and results would not be significantly affected by the small departures from a good fit. The results from human evaluations provide a more practical solution, which reinforces the statements from regression experts that residual plots are an indispensible method for model diagnostics.

>>>>

---
title: |
  A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol
type: ARTICLE TEMPLATE
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a, b, c
    email: emi.tanaka@anu.edu.au
  - name: Susan VanderPlas
    affil: d
    email: susan.vanderplas@unl.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Biological Data Science Institute, Australian National University, Acton, ACT, Australia
  - num: c
    address: |
      Research School of Finance, Actuarial Studies and Statistics, Australian National University, Acton, ACT, Australia
  - num: d
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: paper.bib
abstract: |
  Regression experts consistently recommend plotting residuals for model diagnosis, despite the existence of numerous hypothesis test procedures. This paper provides evidence for why this is good advice, using data from a visual inference experiment. We show how conventional tests are too sensitive to be practically useful, and how the lineup protocol can be used to yield reliable and consistent reading of residual plots for better model diagnosis. 
keywords: |
  statistical graphics; data visualization; visual inference; hypothesis testing; reression analysis; cognitive perception; simulation; practical significance; effect size
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@

```{r different-x-dist-poly-power, fig.pos="t!", fig.height = 6, fig.cap = "Comparison of power on lineups with different fitted value distributions for conventional and visual tests (columns) for non-linearity and heteroskedasticity patterns (rows). The power curves of conventional tests for non-linearity and heteroskedasticity patterns are produced by RESET tests and BP tests, respectively. Power curves of visual tests are estimated using five evaluations on each lineup. For lineups with a uniform fitted value distribution, the five evaluations are repeatedly sampled from the total eleven evaluations to give multiple power curves (grey). Surprisingly, the fitted value distribution has produces more variability in the power of conventional tests than visual tests. Uneven distributions, normal and lognormal distributions, tend to yield lower power.", dev = 'png', dpi = 450}

@@

  # Theme
  theme_light() +
  scale_color_manual(values = rcartocolor::carto_pal(4, "Safe")[c(4,1,3,2)]) +
  xlab(expression(log[e] (Effect_size))) +
  ylab("Power") +
  labs(col = "", linetype = "Test") +
  facet_grid(type~conv_or_not, scales = "free_x") +
  theme(legend.position = "bottom")
```

Figure \ref{fig:different-x-dist-poly-power} examines the impact of the fitted value distribution on the power of conventional (left) and visual (right) tests for both the non-linearity (top) and heteroskedasticity (bottom) patterns. For conventional tests, only the power curves of appropriate tests which are RESET tests and BP tests for non-linearity pattern and heteroskedasticity pattern respectively are shown. For visual tests, note that, more evaluations on lineup with uniform fitted value distribution were collected, so to have a fair comparison, we randomly sample five from the total eleven evaluations to estimate the power curves. Re-sampling produces the multiple curves for the uniform, and provides an indication of the variation of the power.

@@

Motivated by the advice of regression analysis experts, that residual plots as opposed to conventional tests are indispensible methods for assessing model fit, a human subjects experiment was conducted using visual inference. The experiment tested two primary departures from good residuals: non-linearity and heteroskedasticity.

The experiment found that conventional residual-based statistical tests are more sensitive to weak departures from model assumptions than visual tests as would be evaluated by humans. That is, a conventional test conclude there are problems with the model fit almost twice as often as a human would. They often reject when departures in the form of non-linearity and heteroskedasticity are not visibly different from null residual plots.

While it might be argued that the conventional tests are correctly detecting the small but real effects, it can also be considered that the conventional tests are rejecting when it is not necessary. Many of these rejections happen even when downstream analysis and results would not be significantly affected by the small departures from a good fit. The results from human evaluations provide a more practical solution, which reinforces the statements from regression experts that residual plots are an indispensible method for model diagnostics.

### 2023-04-16 paper.tex
switched facetting in Fig 17 to match the previous two figures

\begin{abstract}
Regression experts consistently recommend plotting residuals for model
diagnosis, despite the existence of numerous hypothesis test procedures.
This paper provides evidence for why this is good advice, using data
from a visual inference experiment. We show how conventional tests are
too sensitive to be practically useful, and how the lineup protocol can
be used to yield reliable and consistent reading of residual plots for
better practical application.
\end{abstract}

@@

\caption{Comparison of power on lineups with different fitted value distributions for conventional and visual tests for non-linearity and heteroskedasticity patterns. The power curves of conventional tests for non-linearity and heteroskedasticity patterns are produced by RESET tests and BP tests, respectively. Power curves of visual tests are estimated using five evaluations on each lineup. For lineups with a uniform fitted value distribution, the five evaluations are repeatedly sampled from the total eleven evaluations to give multiple power curves (grey). Surprisingly, the fitted value distribution has produces more variability in the power of conventional tests than visual tests. Uneven distributions, normal and lognormal distributions, tend to yield lower power.}\label{fig:different-x-dist-poly-power}
\end{figure}

Figure \ref{fig:different-x-dist-poly-power} examines the impact of the
fitted value distribution on the power of conventional (top row) and
visual (bottom row) tests for both the non-linearity (left column) and
heteroskedasticity (right column) patterns. For conventional tests, only
the power curves of appropriate tests which are RESET tests and BP tests
for non-linearity pattern and heteroskedasticity pattern respectively
are shown. For visual tests, note that, more evaluations on lineup with
uniform fitted value distribution were collected, so to have a fair
comparison, we randomly sample five from the total eleven evaluations to
estimate the power curves. Re-sampling produces the multiple curves for
the uniform, and provides an indication of the variation of the power.

>>>>

\begin{abstract}
Regression experts consistently recommend plotting residuals for model
diagnosis, despite the existence of numerous hypothesis test procedures.
This paper provides evidence for why this is good advice, using data
from a visual inference experiment. We show how conventional tests are
too sensitive to be practically useful, and how the lineup protocol can
be used to yield reliable and consistent reading of residual plots for
better model diagnosis.
\end{abstract}

@@

\caption{Comparison of power on lineups with different fitted value distributions for conventional and visual tests (columns) for non-linearity and heteroskedasticity patterns (rows). The power curves of conventional tests for non-linearity and heteroskedasticity patterns are produced by RESET tests and BP tests, respectively. Power curves of visual tests are estimated using five evaluations on each lineup. For lineups with a uniform fitted value distribution, the five evaluations are repeatedly sampled from the total eleven evaluations to give multiple power curves (grey). Surprisingly, the fitted value distribution has produces more variability in the power of conventional tests than visual tests. Uneven distributions, normal and lognormal distributions, tend to yield lower power.}\label{fig:different-x-dist-poly-power}
\end{figure}

Figure \ref{fig:different-x-dist-poly-power} examines the impact of the
fitted value distribution on the power of conventional (left) and visual
(right) tests for both the non-linearity (top) and heteroskedasticity
(bottom) patterns. For conventional tests, only the power curves of
appropriate tests which are RESET tests and BP tests for non-linearity
pattern and heteroskedasticity pattern respectively are shown. For
visual tests, note that, more evaluations on lineup with uniform fitted
value distribution were collected, so to have a fair comparison, we
randomly sample five from the total eleven evaluations to estimate the
power curves. Re-sampling produces the multiple curves for the uniform,
and provides an indication of the variation of the power.
